---
title: "SharpVarianceEstimatorVignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SharpVarianceEstimatorVignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sharpvar)
library(ggplot2)
library(latex2exp)
set.seed(123)
```

## [Sharp Bounds on the Variance of General Regression Adjustment in Randomized Experiments](https://arxiv.org/pdf/2411.00191)

In this vignette, we are going to illustrate how to use our sharp variance bound package and illustrate the benefits of sharp variance bounds more generally.

## 1) Simple Showcase of Sharp Variance Bounds

In this section, we simulate data that showcases the benefits of sharp variance bounds. The simulation is outlined in Section 4. of ([Mikhaeil and Green, 2024](https://arxiv.org/pdf/2411.00191)).

```{r}

population <-sharpvar::create_dataset(1000,0.5,1)
Y1 <- population$Y1
Y0 <- population$Y0
x <- population$X

dat <- sharpvar::randomize(Y0,Y1,x,0.5)
mod <- lm(Y_obs ~ treatment*I(X-mean(X)) , data=dat)


conventional_var <- sharpvar::neyman_conventional(mod$residuals,dat$treatment)
cs_var <- sharpvar::neyman_cs(mod$residuals,dat$treatment)
sharp_var <-sharpvar::sharpvar(mod$residuals,dat$treatment,upper=TRUE)

conv_cor <- sharpvar::cor_approx_conventional(mod$residuals,dat$treatment)
sharp_cor <-sharpvar::cor_approx_sharp(mod$residuals,dat$treatment)
cs_cor <- sharpvar::cor_approx_cs(mod$residuals,dat$treatment)
 
```

```{r}
print(sharp_var/cs_var)
print(sharp_var/conventional_var)

```

## 2) Sweeps

This section recreates the plots in Section 4 of ([Mikhaeil and Green, 2024](https://arxiv.org/pdf/2411.00191)).

### 2a) Sweep over simulation parameter

Sweeping over the simulation parameter allows to trace out the operation characteristics of sharp variance bounds under different levels of heterogeneity. The extremes $\theta = 0$ and $\theta =1$ correspond to the sharp null and linear dependence between the adjusted potential outcomes, respectively.

```{r}
cs_var <- c()
conventional_var <- c()
true_te <- c()
te_est <- c()
sharp <- c()
sharp_cor <- c()
ney_cor <- c()
cs_cor <-c()
true_cor <- c()
r2 <- c()
sweep <-seq(0,1,0.2) 
for (sd in sweep){
population <-sharpvar::create_dataset(2000,sd)
Y1 <- population$Y1
Y0 <- population$Y0
x <- population$X

true_te <-c(true_te,mean(Y1) - mean(Y0))
mod_1 <- lm(Y1 ~ I(x-mean(x)))
mod_0 <- lm(Y0 ~ I(x-mean(x)))
b1 <- coef(mod_1)["I(x - mean(x))"]
b0 <- coef(mod_0)["I(x - mean(x))"]

true_cor <- c(true_cor,cov(Y1-b1*(x-mean(x)),Y0-b0*(x-mean(x))))

rep <- vapply(1:1000,function(i){  #100000
  dat <- sharpvar::randomize(Y0,Y1,x,0.5)
  mod <- lm(Y_obs ~ treatment*I(X-mean(X)) , data=dat)

  r2_rand<-summary(mod)$r.squared
  te_rand <- coef(mod)["treatment"]
  var_conv <- sharpvar::neyman_conventional(mod$residuals,dat$treatment)
  var_cs <- sharpvar::neyman_cs(mod$residuals,dat$treatment)
  sharp_rand <-sharpvar::sharpvar(mod$residuals,dat$treatment)
  
  sharp_cor_r <- sharpvar::cor_approx_sharp(mod$residuals,dat$treatment)
  ney_cor_r <- sharpvar::cor_approx_conventional(mod$residuals,dat$treatment)
  cs_cor_r <- sharpvar::cor_approx_cs(mod$residuals,dat$treatment)
  

  list <- c(r2_rand,te_rand, var_conv , var_cs,sharp_rand, sharp_cor_r , ney_cor_r ,  cs_cor_r)
},numeric(8))

r2 <- c(r2,mean(rep[1,]))
cs_var <- c(cs_var, mean(rep[4,]))
conventional_var <- c(conventional_var, mean(rep[3,]))

sharp <- c(sharp,mean(rep[5,]))
te_est <- c(te_est,mean(rep[2,]))
sharp_cor <- c(sharp_cor,mean(rep[6,]))
ney_cor <- c(ney_cor,mean(rep[7,]))
cs_cor <- c(cs_cor,mean(rep[8,]))
}
sd_sweep<- data.frame(s = sweep, sharp= sharp, cs_var = cs_var ,conventional_var = conventional_var,sharp_cor = sharp_cor,ney_cor = ney_cor, cs_cor = cs_cor , true_cor = true_cor, te_est = te_est,true_te = true_te)
```

```{r}

ggplot(data=sd_sweep)+
  
    geom_line(aes(x=s,y=cs_cor),color="skyblue")+

    geom_line(aes(x=s,y=true_cor),color="black")+
    geom_line(aes(x=s,y=ney_cor),color="dodgerblue4")+
    geom_line(aes(x=s,y=sharp_cor),color="cornflowerblue")+

  theme_minimal()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))+
ylab(latex2exp::TeX("Estimate of the Cross Term")) +
xlab(latex2exp::TeX("Simulation Parameter $\\theta$" ))+ theme(legend.position = "none")+
annotate(geom="text", x=0.5, y=12.5, label="Conventional",color="dodgerblue4")+
annotate(geom="text", x=0.5, y=6.5, label="Cauchy-Schwarz",color="skyblue")+
annotate(geom="text", x=0.5, y=4, label="Sharp",color="cornflowerblue")+
annotate(geom="text", x=0.5, y=1.8, label="True",color="black")+
scale_x_continuous(expand = c(0,0.02),limits=c(0,1))+
scale_y_continuous(expand = c(0,0.02),limits=c(0,15))

```

### 2b) Sweep over N

In this section, we compare the asymptotics of the bias for conventional, Cauchy-Schwarz and Sharp Variance estimator under the sharp null.

```{r}
sweep_over_N <- function(N_up,noise_level){
cs_var <- c()
conventional_var <- c()
true_var <- c()
sharp <- c()
r2 <- c()
sweep <-seq(10,N_up,50) #10
population <-sharpvar::create_dataset(N_up,0,noise_level)
for (N in sweep){
Y1 <-  population$Y1[1:N]
Y0 <- population$Y0[1:N]
x <- population$X[1:N]
mod_1 <- lm(Y1 ~ I(x-mean(x)))
mod_0 <- lm(Y0 ~ I(x-mean(x)))
b1 <- coef(mod_1)["I(x - mean(x))"]
b0 <- coef(mod_0)["I(x - mean(x))"]
true_var <- c(true_var,sharpvar::var_true(mod_1$residuals,mod_0$residuals))
rep <- vapply(1:1000,function(i){  #100000
dat <- sharpvar::randomize(Y0,Y1,x,0.5)
mod <- lm(Y_obs ~ treatment*I(X-mean(X)) , data=dat)
r2_rand<-summary(mod)$r.squared
var_conv <- sharpvar::neyman_conventional(mod$residuals,dat$treatment)
var_cs <- sharpvar::neyman_cs(mod$residuals,dat$treatment)
sharp_rand <-sharpvar::sharpvar(mod$residuals,dat$treatment)
c(r2_rand, var_conv , var_cs,sharp_rand)
},numeric(4))
r2 <- c(r2,mean(rep[1,]))
conventional_var <- c(conventional_var, mean(rep[2,]))
cs_var <- c(cs_var, mean(rep[3,]))
sharp <- c(sharp,mean(rep[4,]))
}
return(data.frame(s = sweep, sharp= sharp, cs_var = cs_var ,conventional_var = conventional_var, true = true_var,r2=r2,noise=noise_level))}
```

```{r}
d_1 <- sweep_over_N(600,1)
d_10 <- sweep_over_N(600,10)

```

```{r}
ggplot(data=rbind(d_1,d_10))+
    geom_line(aes(x=s,y=abs(conventional_var-true),group=noise),color="dodgerblue4")+
    geom_line(aes(x=s,y=abs(cs_var-true),group=noise),color="skyblue")+
    geom_line(aes(x=s,y=abs(sharp-true),group=noise),color="cornflowerblue")+
    
  theme_minimal()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))+
ylab(latex2exp::TeX("Log(Bias)")) +
xlab(latex2exp::TeX("Sample Size $N$" ))+ theme(legend.position = "none")+
annotate(geom="text", x=200, y=0.2, label=latex2exp::TeX("$R^2 = 0.03$" ),color="black")+
annotate(geom="text", x=200, y=0.001, label=latex2exp::TeX("$R^2 = 0.35$" ),color="black")+

scale_y_continuous(expand = c(0,0.02),trans="log",labels = scales::number_format(accuracy = 0.0001,decimal.mark = '.'),breaks=c(0.0001,0.001,0.01,0.1,1))+
  
scale_x_continuous(expand = c(0,.1),limit=c(0,510))

```
